---
title: "Practical"
author: "James Mburu"
date: "February 17, 2016"
output: beamer_presentation
header-includes:  
- \usetheme{CambridgeUS}
- \useinnertheme{rounded}
- \usecolortheme{sidebartab}
- \usefonttheme{professionalfonts}
- \usepackage{xcolor,soul,hyperref,cite,graphicx,amsmath,verbatim,longtable}
- \hypersetup{colorlinks,urlcolor=blue}
linkcolor: blue
urlcolor: blue
---


```{r library,include=FALSE,warning=FALSE}

lib <- c("epiDisplay", "plyr" , "base", "foreach", "foreign","ggplot2",
         "grid","dplyr","tidyr","grid","gridExtra","xtable")
lapply(lib, library, character.only = TRUE)

```

## 


\begin{center}

\textbf{Graphical Exploration}\\      


\end{center}


## Exercise. 1

Draw a box plot to show the distribution of the birth weight(bweight) for the different age groups (agegrp).        

-. Label the titles i.e y, x, and the main title      

-. Change the color of the titles from the default black to a color of your choice.

-. Increase the size of axis titles and  main title

-. save the graph as a .pdf file to a location of your choice        


## Exercise. 2

Draw a Grouped bar plot showing the number of kids in each category of lbw grouping them by sex.      


-. Label the titles i.e y, x, and the main title      

-. Change the color of the titles from the default black to a color of your choice.

-. Increase the size of axis titles and main title.

-. save the graph as a .png file to a location of your choice    

## Solution 1

\small
```{r,include=FALSE}
birth<-read.dta("H:/Jmburu/Class/Data/birthweight2.dta")
```


\small
```{r,comment=NA}
bx<-ggplot(birth, aes(x=agegrp, y=bweight, color=agegrp)) +
    geom_boxplot() + guides(color=FALSE)
bx<- bx + theme_bw() + labs(title="Birth weight per age group"
                            ,x="",y="Birth weight(g)")
bx<- bx + theme(title=element_text(color="blue", size=18),
  axis.title.y=element_text(size=15,face="bold",color="brown"),
  axis.text=element_text(size=14) )
```

```{r,eval=FALSE}
# Save the plot
ggsave("H:/Jmburu/boxplot.pdf",plot=bx,width=8,height=6,unit="in")

```

## Solution 1 ... 
 
```{r,echo=FALSE,fig.height=6,fig.width=9}
bx
```


## Solution 2 ... 

\small

```{r,include=FALSE}
(Count<-table(birth$lbw,birth$sex))
Count<- data.frame(Count)
names(Count)<- c("lbw","sex","count")
```


```{r}
b<-ggplot(data=Count, aes(x=sex,y=count,fill=lbw))
b<- b + geom_bar(stat="identity",width=.5,position ="dodge") +
    theme_bw()
b<- b + labs(title="Grouped Bar Plot",x="Sex",y="Counts") 
b<- b + theme(legend.key=element_blank(),
           legend.title=element_blank(),
           legend.text=element_text(size = rel(1.1)),
           legend.direction="horizontal",legend.position="bottom")
```

```{r,eval=FALSE}
# Save the plot
ggsave("H:/Jmburu/barplot.png",plot=b,width=8,height=6,unit="in")

```

## Solution 2 ...

```{r,echo=FALSE,fig.height=6,fig.width=9}
b
```

 
